---
title: "Latent Growth Curve Modeling (LGCM)"

output:
  html_document:
    css: "custom-lsem.scss"
    df_print: kable
    code_folding: hide
    number_sections: no
    mathjax: default
    theme: flatly
    toc: true
    toc_collapsed: true
    toc_float: true
    toc_depth: 3
    includes:
      before_body: [include_lsem_nav.html]
      after_body: [include_lsem_nav_end.html, footer.html]
      
---

<!-- 
split_by: section
split_by: chapter
bookdown::gitbook
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE, echo = FALSE}
# packages
library(tidyverse) # xxxxxx
library(lavaan)
library(haven)
library(readxl)
library(DiagrammeR)
library(knitr)
library(pander)
library(report)
library(equatiomatic)
library(stargazer)
```

```{r, echo = FALSE}
# Load the data:
mydata <- read_sav("/Users/shawes/Desktop/Biostatistics/data/TestData.sav")
```

<br />

The latent growth model (LGM) extends CFA to the analysis of items across time.
The LGCM approach provides a flexible strategy for specifying and evaluating relationships among variables that have been assessed repeatedly across multiple measurement occasions [@Preacher2008]. These models are often the go-to choice for developmental research as they allow for characterizing the rate (e.g., increasing, decreasing) and shape (e.g., linear, quadratic) of change in outcomes at both, the group- and individual-level. The LGCM framework can also be extended to evaluate other substantive questions relevant to continuity and change. This vignette will explore a basic ("unconditional") LGCM, as well as three commonly applied extensions to this approach, as follow:

- [5.1: LGCM w/ Continuous Indicators](lgcm-ex-1.html)
- [5.2: LGCM w/ Continuous Indicators with Time-Varying Covariates](lgcm-ex-1.html)
- [5.3: LGCM w/ Continuous Indicators Predicting Distal Outcomes](lgcm-ex-1.html)
- [5.4: LGCM w/ Continuous Indicators and Antecedent Predictors](lgcm-ex-1.html)
- [5.5: xxxxxxxxx other examples See statmodel Chpt. 6 sections](lgcm-ex-1.html)
<br /> <br />

## 5.1: Basic LGCM {.tabset .tabset-fade style="color: black;"}

This is an example of a basic LGCM w/ continuous outcome [y], across four measurement occassions

<br />

### <span style="color:red">Text Overview</span>

LGCM is a special class of confirmatory factor analysis used to model some outcome (Y) measured repeatedly over time (t). The most basic LGCM includes two latent factors and uses a fixed time structure to model the observed outcome (Y) across assessment waves (t): 1) a latent intercept; and 2) a latent slope. The intercept factor is constant for each individual across observations, and is akin to the point at which a straight line "intercepts" the vertical axis on a two-dimensional coordinate system (@citation). This is modeled by fixing the factor loadings for the latent intercept to 1 at each measurement occassion. 

### <span style="color:red">Diagram</span>

```{r growth_graph, echo=FALSE}
grViz('fig/growth.gv', width='100%', height='25%')
fig.cap = "test test test test"
```

### <span style="color:blue">Code Explanation</span>

```{r Example_5-1, eval=FALSE, class.source = 'fold-show'}
# Example-5.1
model <- ' i =~ 1*y1 + 1*y2 + 1*y3 + 1*y4
           s =~ 0*y1 + 1*y2 + 2*y3 + 3*y4 '
fit <- growth(model, data=path-to-data)
summary(fit)
```

A basic linear growth model includes two factors: 1) a latent intercept; and 2) a latent slope. This model estimates the following six parameters:

  1) intercept mean ($\[ \mu\]$i); 
  2) intercept variance ($\[ \sigma\]$i); 
  3) slope mean ($\[ \mu\]$s); 
  4) slope variance ($\[ \sigma\]$s); 
  5) intercept and slope covariance ($\[ \sigma\]$is); and 
  6) an error variance that remains constant across repeated measurements ($\[ \sigma\]$ε)
<br /> <br />

This model is specified by using the '*=~*' operator to name and define the growth model factors. The left-hand side of the '=~' operator is used to specify the latent intercept *i* and latent slope *s* factors. The right-hand side of the '=~' operator specifies the growth model outcome and time scores. In the LGCM framework, the functional form of growth (e.g., linear) is defined by fixing the  factor loadings of the growth model factors. The intercept factor remains constant for each individual across time; as such, factor loadings for the intercept *i* are fixed to the value of 1 at each measurement occassion. To define linear growth across the 4 equidistant time points, time scores for the slope *s* growth factor are fixed at 0, 1, 2, and 3. Setting the time score at the first measurement occassion to zero anchors the latent intercept as "baseline" or "initial status". Means and variances of the growth factors are estimated as the default, as is the growth factor covariance.

```{r, eval=FALSE, class.source = 'fold-show'}
model <- ' i =~ 1*y1 + 1*y2 + 1*y3 + 1*y4
           s =~ 0*y1 + 1*y2 + 2*y3 + 3*y4 '
```

The *growth()* function can be called to fit the *model* that we defined in the prior step.

```{r, eval=FALSE, class.source = 'fold-show'}
fit <- growth(model, data=path-to-data)
```

The *summary()* function can be called to provide an overview and show results of the fitted model.

```{r, eval=FALSE, class.source = 'fold-show'}
summary(fit)
```

<!--

The growth() function, is a wrapper function for the lavaan() function that automatically specifies fixed and free parameters for the latent growth model. Use of this function eliminates several additional coding steps such as having to constrain a subset of observed intercepts to zero and specifying that latent means are to be freely estimated. See [Example 5 full code] for additional details. 

fitMeasures

The *fitmeasures()* function can be called to provide a variety of model fit indices.

-->

### <span style="color:green">ABCD Example</span>

"The goal of our current LGM is to assess the linear trajectory of GPA, with the question being, does the average student GPA in a particular school go up or down over time?"

```{r}
# LGCM for a continuous outcome [y], across four mesurement occassions
model <- ' i =~ 1*Brain_T1 + 1*Brain_T2 + 1*Brain_T3 + 1*Brain_T4
           s =~ 0*Brain_T1 + 1*Brain_T2 + 2*Brain_T3 + 3*Brain_T4 '
fit <- growth(model, data=mydata, meanstructure=TRUE, estimator = "MLR", missing = "ML")
summary(fit, standardized = TRUE, fit.measures = TRUE)

# extract information
mean_lv <- lavInspect(fit, "mean.lv", add.labels = TRUE, add.class = TRUE,
           list.by.group = TRUE,
           drop.list.single.group = TRUE)

cov_lv <- lavInspect(fit, "cov.lv", add.labels = TRUE, add.class = TRUE,
           list.by.group = TRUE,
           drop.list.single.group = TRUE)


intercept_var <- cov_lv[1]
slope_var <- cov_lv[4]
is_covary <- cov_lv[2]

intercept_var
slope_var
is_covary

intercept_mean <- mean_lv[1]
slope_mean <- mean_lv[2]

#lavInspect(fit, "sampstat")
#lavNames(fit, "ov" "lv")
#lavInspect(fit, "mean.lv") #Alias: "eeta" [for E(eta)].
#lavInspect(fit, "cor.lv")
#parTable(fit)
#varTable(fit)
#inspect(fit, what="cor.all")
#lavCor(fit)
#out <- lavExport(fit, target = "Mplus", export=FALSE)
#cat(out)

```

```{}

Model Test User Model:
##                                                Standard      Robust
##   Test Statistic                                392.563     312.386
##   Degrees of freedom                                  5           5
##   P-value (Chi-square)                            0.000       0.000

```

```{}
User Model versus Baseline Model:

Comparative Fit Index (CFI)                    0.848       0.789
Tucker-Lewis Index (TLI)                       0.818       0.747

```

```{}
Root Mean Square Error of Approximation:
 
   RMSEA                                          0.242       0.215

```

```{}
Standardized Root Mean Square Residual:
 
   SRMR                                           0.238       0.238
```






<!-- link to model diagnostics to evaluate in instances when fit is poor --> The average baseline score (i.e., the mean intercept value) on the xxxxx measure is `r intercept_mean` and the mean slope value is `r slope_mean`. This means that the predicted xxxx score at the first measurement occasion (i.e., t=0) is `r intercept_mean` and at each subsequent measurement occasion the xxxx score is expected to go up by `r slope_mean` points. There is significant variability in individual's xxxxx scores at baseline `r intercept_var`, p < .xxx), as well as in their slope value `r slope_var`, p < .001), suggesting that the rate of change in xxxx was not the same for all children. Finally, the estimated negative covariance or `r is_covary`, p < .001) implies that the higher the starting score on xxxx, the weaker the increase in xxxxxx.

#### Plots

```{r wide-to-long, echo = FALSE}
new <- mydata %>%
  select(id_s, sex, Brain_T1, Brain_T2, Brain_T3, Brain_T4)

new2 <- new %>% 
  pivot_longer(!id_s:sex, values_drop_na = TRUE) %>% 
  mutate(checked = TRUE)

## change id and male to factor variables
#new2 <- within(new2, {
#    sex <- factor(sex, levels = 0:1, labels = c("female", "male"))
#    name <- factor(name, levels = "Brain_T1" "Brain_T2" "Brain_T3" "Brain_T4", labels = c("T1", "T2", "T3", "T4"))
#})

#head(new2)
```

```{r growth model plots}

p <- ggplot(data = new2, aes(x = name, y = value, group = id_s))

p + geom_line() + stat_smooth(aes(group = 1), method = "lm", se = FALSE) + stat_summary(aes(group = 1), geom = "point", fun = mean, shape = 17, size = 3)+theme_bw()
#+ facet_grid(. ~ sex)
#+ facet_wrap(~id_s)

ggplot(new2, aes(x=name, y=value, group=1))+geom_jitter(height = 0, width = 0.2, alpha = 0.2)+geom_point(alpha=0.5)+stat_summary(fun=mean, colour="red", geom="line", size = 3)+theme_bw() # draw a mean line in the data

#shape = male, 
#linetype = male
#geom_line(aes(linetype = male), size = 1)

## load the grid package for the unit function to adjust the legend width
#require(grid)

#theme(legend.key.width = unit(1, "cm")) + theme_bw() + scale_linetype(name = "Sex"


```

#### Tables
```{r}
head(mydata)
```

Start with the 'intercepts':

```{}
Intercepts:
                   Estimate  Std.Err  Z-value  P(>|z|)
                       
    i                 0.487    0.048   10.072    0.000
    s                 0.267    0.045    5.884    0.000
```


```{}
Covariances:
                   Estimate  Std.Err  Z-value  P(>|z|)
  i ~~                                                
    s                 0.226    0.050    4.512    0.000
Variances:
                   Estimate  Std.Err  Z-value  P(>|z|)
    outcome0          0.287    0.041    6.924    0.000
    outcome1          0.219    0.021   10.501    0.000
    outcome2          0.185    0.027    6.748    0.000
    outcome3          0.357    0.065    5.485    0.000
    i                 0.977    0.076   12.882    0.000
    s                 0.969    0.065   14.841    0.000
```

## {-}
<br /> <br />

## 5.2: LGCM w/ Time-Varying Covariates {.tabset .tabset-fade style="color: black;"}
<br />
test

### <span style="color:orange">Brief Overview</span>
test

### <span style="color:red">Diagram</span>
```{r, echo=FALSE}
library(DiagrammeR)

grViz("
digraph narc {

  # a 'graph' statement
  graph [overlap = true, fontsize = 12]

  # nodes for observed and latent
  node [shape = circle, fontname = Helvetica]
  'Intercept'; 'Slope';
  
  node [shape = box, fontname = Helvetica]
  t1; t2; t3; t4;

  # edges
  'Intercept'->'t1' 'Intercept'->'t2' 'Intercept'->'t3' 'Intercept'->'t4'
  'Slope'->'t1' 'Slope'->'t2' 'Slope'->'t3' 'Slope'->'t4'
}")
```

### <span style="color:blue">Code</span>
```{r}
# a linear growth model with a time-varying covariate
model <- '
  # intercept and slope with fixed coefficients
    i =~ 1*Brain_T1 + 1*Brain_T2 + 1*Brain_T3 + 1*Brain_T4
    s =~ 0*Brain_T1 + 1*Brain_T2 + 2*Brain_T3 + 3*Brain_T4
  # regressions
    i ~ sex + eduPar
    s ~ sex + eduPar
  # time-varying covariates
    Brain_T1 ~ G_CD
    Brain_T2 ~ IQ
    Brain_T3 ~ age2
    Brain_T4 ~ White
'
fit <- growth(model, data = mydata)
summary(fit)
```

### <span style="color:green">Code Commentary</span>
test test test

## {-}
<br /> <br />

---

## 5.3: LGCM Antecedent Predictors {.tabset .tabset-fade style="color: black;"}
<br />
test test test

### <span style="color:orange">Brief Overview</span>
### <span style="color:red">Diagram</span>
### <span style="color:blue">Code</span>
### <span style="color:green">Code Commentary</span>
## {-}
<br /> <br />

---

## 5.4: LGCM Predicting a Distal Outcome {.tabset .tabset-fade style="color: black;"}
<br />
test test test

The prior section introduced the fundamental latent growth model with an intercept and linear growth factor. This model sufficiently addresses the question of whether the linear trajectory is increasing, decreasing or flat over time. In our example, we found that GPA increases by  on average for every one semester increase in time. Other researchers may be interested in finding casual (using the word loosely) predictors of this linear trajectory. For example, we found that the trajectory of GPA increases over time, but does gender predict this trend? More specifically, are there gender differences in either the starting GPA (intercept factor) or the growth trajectory (slope factor)? To find out, we would need to include an observed predictor of the latent mean and slope terms.

### <span style="color:orange">Brief Overview</span>
### <span style="color:red">Diagram</span>
### <span style="color:blue">Code</span>
### <span style="color:green">Code Commentary</span>

---
***
## [Continue to Latent Transition Analysis (LTA)](lsem-4.html){.continue-link}
